Parsing Postfix Log Files

### 2009-04-26 doc/logparser-complications.tex
Changes made after proofreading the preface and chapters 1-3.
Changes based on proofreading chapters 4 and 5.

    \item [Lack of information in the log files] Seven complications are
        caused by lack of information in the log files.  In general, some
        extra logging by Postfix could remove these complications, and for
        one complication, extra logging in later versions of Postfix does
        remove it (\sectionref{identifying bounce notifications}); three of
        the complications would be removed if Postfix generated a log line
        whenever it discarded a mail.

@@

A delivery attempt lacks a queueid until one recipient has been accepted,
so log lines must first be correlated by \daemon{smtpd} \acronym{pid}, then
transition to being correlated by their queueid.  This is relatively minor,
but does require:

@@

    \item Each action needs to know whether it deals with accepted mails,
        and should find the mail in the by\_queueid state tables, or deals
        with delivery attempts, and thus will find the connection in the
        by\_pid state tables.  Some actions, e.g.\
        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
        functions fall into two groups: those that deal with state tables,
        and so need to know which one to operate on; the functions in the
        other group are passed a specific mail or connection to work with,
        and do not need to know about state tables at all.

@@

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses because of aliasing.
When this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently, the
parent mail will log the creation of the child mail, e.g.\ parent mail
\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:

@@

Because of the uncertain order that log lines can appear in, \parsername{}
cannot complain when it encounters a log line from either \daemon{qmgr} or
\daemon{cleanup} for a mail that does not exist in the state tables;
instead it must create the state table entry and flag the mail as coming
from an unknown origin.  Subsequently invoked actions will clear that flag
if the origin of the mail becomes clear.  Obviously the parser could omit
checking where mails originate from, but requiring an explicit source helps
to expose bugs in the parser; such checks helped to identify the
complications described in \sectionref{discarding cleanup log lines} and
\sectionref{pickup logging after cleanup}.

Process scheduling can have a still more confusing effect: the child mail
will often be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  Unfortunately, it is not possible
to distinguish between child mails waiting to be tracked and other mails
whose origin is unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} cannot be expected to fully understand and reconstruct a mail
when it is lacking some of that mail's log lines.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; the lof file will have log
lines for a mail that lacks an obvious source.  Similarities exist to the
problem of identifying re-injected mails discussed in \sectionref{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line that identifies their
source.  Heuristics must be used to identify bounce notifications:

@@

    \item The message-id must have the correct format and correspond with
        the queueid of the mail.  Although a mail sent from elsewhere could
        easily have the correct message-id format, the chance that the
        queueid in the message-id would correspond with the queueid of the
        mail is extremely small.

@@

If a mail is misclassified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably misclassification of
this kind is a benefit rather than a drawback, because other mails
generated internally by Postfix will be handled correctly.  Postfix 2.3 and
subsequent versions log the creation of a bounce message.

@@

Postfix does not log a message making the client's behaviour clear, so
again heuristics are required to identify when a delivery attempt is
aborted in this way.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each accepted mail is
checked for the following:

@@

    \item If third and subsequent results exist, were those log lines
        processed by the \action{SAVE\_DATA} action?  Any log lines after
        the first two they should be informational only.

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is removed form the state tables.  There will
be no further entries logged for such mails, so without identifying and
discarding them they accumulate in the state table and will cause clashes
if the queueid is reused.  The mail cannot be entered in the database
because the only data available is the client hostname and IP address, but
the database schema requires many more fields be populated (see
\sectionref{connections table} and \sectionref{results table}).  These
heuristics are quite restrictive, and appear to have little scope for false
positives; any false positives would cause a warning when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser, e.g.\ \sectionref{timeouts during data phase}.  Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection log line and a mail acceptance log
        line, followed by one or more delivery rejected log lines.

@@

If both checks are successful then the mail is assumed to have been
discarded by Postfix when the client disconnected; \parsername{} will
remove it from the state tables.  There will be no further log lines for
such mails, so if \parsername{} does not deal with them immediately they
accumulate in the state table and will cause clashes if the queueid is
reused.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes a timeout occurs or the
connection is lost during the DATA phase;\footnote{For the sake of brevity
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and \parsername{} needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs no data will be available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the \action{TIMEOUT} action does the following:

@@

    \item Record the timeout and captured data in the connection's results.

    \item If no mails have been accepted yet nothing needs to be done; the
        \action{TIMEOUT} action ends.

@@

            \item The timestamp of the log line preceding the timeout log
                line is earlier than the timestamp of the last accepted
                delivery attempt, i.e.\ there have not been any rejections
                since then the delivery attempt was accepted.

@@

            \item The mail does not have a \daemon{qmgr} log line.

@@

        If all the checks pass the mail will be discarded from the state
        tables, and will not be entered in the database.  If one or more
        checks do not pass it is assumed that the timeout happened after a
        rejected delivery attempt.  This assumption is not necessarily
        correct, because Postfix could have accepted an earlier recipient,
        rejected a later one, and continued to accept delivery of the mail
        for the first recipient.  In that case the timeout applies to the
        partially accepted mail, which should be discarded; however, this
        has not occurred in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail
because it will not have enough data to satisfy the database schema (see
\sectionref{timeouts during data phase}); parsing the \daemon{cleanup} log
line causes the \action{CLEANUP\_PROCESSING} action to create a new state
table entry, to help deal with re-injected mails (\sectionref{Re-injected
mails}).  This is incorrect because the log line actually belongs to the
mail that has just been discarded; if the queueid is reused there will be a
queueid clash, otherwise the new mail will just remain in the state tables.

During the \action{TIMEOUT} action, if the \daemon{cleanup} log line is
still pending, the action updates the \texttt{timeout\_queueids} state
table, adding the queueid and the timestamp from the log line.  The
following checks will be performed for each \daemon{cleanup} log line
that is processed:

@@

    \item Does the \texttt{timeout\_queueids} state table have an entry for
        the queueid in the log line?  If an entry is found it will be
        removed, regardless of whether the remaining criteria are
        satisfied.

    \item Has the queueid been reused yet, i.e.\ does it have an entry in
        the state tables?

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
\action{CLEANUP\_PROCESSING} action blindly discarded the next
\daemon{cleanup} log line after a timeout it would sometimes be mistaken.

@@

When mail is submitted locally, \daemon{pickup} processes the new mail and
generates a log line.  Occasionally this log line will occur later in the
log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
will find that a state table entry already exists for that queueid.
Normally when this happens a warning is generated by the \action{PICKUP}
action, but if the following conditions are met it is assumed that the log
lines were out of order:

@@

    \item The only program that has logged anything thus far for the mail
        is \daemon{cleanup}.

@@

As always with heuristics there may be circumstances in which these
heuristics match incorrectly, but none have been identified so far.  This
complication seems to occur during periods of particularly heavy load, so
is most likely caused by process scheduling vagaries.

@@

    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
        connection must be cleaned up, entered in the database if it has
        enough data, and deleted from the connections state table.

    \item Every Postfix process uses a watchdog which kills the process if
        it is not reset for a considerable period of time (five hours by
        default).  This safeguard prevents Postfix processes from running
        indefinitely and consuming resources if a failure causes them to
        enter a stuck state.

@@

The circumstances above account for all occasions where a \daemon{smtpd}
suddenly stops logging.  In addition to removing an active connection the
last accepted mail may need to be discarded, as detailed in
\sectionref{timeouts during data phase}; otherwise the queueid state table
is untouched.

@@

Occasionally a log file will have out of order log lines that cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
logging after cleanup}.  In the \numberOFlogFILES{} log files used for
testing this problem occurs only five times in 60,721,709 log lines.  All
five occurrences have the same characteristics: the \daemon{local} log line
showing delivery to a local mailbox comes after the \daemon{qmgr} log line
showing removal of the mail from the queue because delivery is complete.
This causes problems: the data in the state tables for the mail is not
complete, so it cannot be entered into the database; a new mail is created
when the \daemon{local} log line is processed and remains in the state
tables; four warnings are issued per pair of out of order log lines.

The \action{COMMIT} action examines the list of programs that have produced
log lines for each mail, checking the list against a table of known-good
Postfix component combinations.  If the mail's combination is found in the
table it can be entered in the database; if the combination is not found
entry must be postponed and the mail flagged for later entry.  The
\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
lines have caused the mail to reach a valid combination then its entry in
the database will proceed, otherwise it must be postponed once more.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but those described in this section only occur
four times in the \numberOFlogFILES{} log files used for testing.  The
symptoms are the same as in \sectionref{aborted delivery attempts}, except
that the mail \textit{has\/} a \daemon{cleanup} log line; nothing can be
found in the log files to explain why this mail does not have any further
log lines.  The only way to detect these mails is to periodically scan all
mails in the state tables, deleting any mails with the following
characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted \parsername{} cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail before Postfix has had a chance to attempt delivery.  The
recipient address will not have been logged yet, and the deleted mails will
not have any more log lines for it to be logged in.  A dummy recipient
address needs to be added, because every mail is required by the database
schema (\sectionref{results table}) to have at least one recipient.
Typically when this complication occurs the log file will show many
instances of it, closely grouped.

This lack of information cannot easily be overcome: it is trivial to log a
warning for every accepted recipient, but Postfix will not yet have
allocated a queueid for the mail when the warning for the first recipient
is logged, so the warning will be associated with the connection rather
than the accepted mail.  A queueid will be allocated after Postfix accepts
the MAIL FROM command if \texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is
set to ``no'', but that setting will cause disk IO for every delivery
attempt, instead of just for delivery attempts where recipients are
accepted, and consequently a drastic reduction in the performance of the
mail server.

@@

\subsection{Bounce Notification Mails Are Delivered Before Their Creation
Is Logged}

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication, bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their source.  To deal with
this the \action{COMMIT} action maintains a cache of recently committed
bounce notification mails, which the \action{BOUNCE\_CREATED} action
subsequently checks if the bounce mail is not found in the state tables.
If the queueid exists in the cache, and its start time is less than ten
seconds before the timestamp of the bounce log line, it is assumed that the
bounce notification mail has already been processed and the
\action{BOUNCE\_CREATED} action does not create one.  If the queueid exists
in the cache it is removed, because it has either just been used or the
problem did not occur for that mail.  Whether the \action{BOUNCE\_CREATED}
action creates a mail or finds an existing mail in the state tables, it
flags the mail as having been seen by the \action{BOUNCE\_CREATED} action;
if this flag is present the \action{COMMIT} action will not add the mail to
the cache of recent bounce notification mails.  This is not required to
correctly deal with the complication, but is an optimisation to reduce the
parser's memory usage.  On the occasions the author has observed this
complication occurring there have been a huge number of bounce notification
mails generated: if every bounce notification mail was cached it would
dramatically increase the memory requirements of the parser.  The cache of
bounce notification mails will be pruned whenever the parser's state is
saved, though if the size of the cache ever becomes a problem it could be
pruned periodically to keep it in check.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{smtp}, or \daemon{virtual}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The \action{DELETE} action adds deleted mails to a cache,
which is checked by the \action{SAVE\_DATA} action, and the current log
line discarded if the following conditions are met:

@@

Sadly this solution involves discarding some data, but the complication
only arises eight times in quick succession in one log file, which is not
in the \numberOFlogFILES{} log files used for evaluating the parser; if
this complication occurred more frequently it might be desirable to find
the mail in the database and add the log line to it.

>>>>

    \item [Lack of information in the log files] Seven complications are
        caused by lack of information in the log files.  In general, some
        extra logging by Postfix could remove these complications, and for
        one complication (\sectionref{identifying bounce notifications}),
        extra logging in later versions of Postfix does remove it; three of
        the complications would be removed if Postfix generated a log line
        whenever it discarded a mail.

@@

A delivery attempt lacks a queueid until the first recipient has been
accepted, so log lines must first be correlated by \daemon{smtpd}
\acronym{pid}, then transition to being correlated by their queueid.  This
is relatively minor, but does require:

@@

    \item Each action needs to know whether it deals with accepted mails,
        and should find the mail in the \texttt{queueid} state table, or
        deals with delivery attempts, and thus will find the connection in
        the \texttt{connections} state table.  Some actions, e.g.\
        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
        functions fall into two groups: one group deals with state tables,
        and so need to know which one to operate on; the functions in the
        other group are passed a specific mail or connection to work with,
        and do not need to know about state tables at all.

@@

The \action{CLONE} action is responsible for copying an entry from the
\texttt{connections} state table to the \texttt{queueids} state table.

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses because of aliasing.
When this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail does not have an origin.  Subsequently, the parent
mail will log the creation of the child mail, e.g.\ parent mail
\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:

@@

Because of the uncertain order that log lines can appear in, \parsername{}
cannot complain when it encounters a log line from either \daemon{qmgr} or
\daemon{cleanup} for a mail that does not exist in the state tables;
instead it must create the state table entry and flag the mail as coming
from an unknown origin.  Subsequently invoked actions will clear that flag
if the origin of the mail becomes clear.  The parser could omit checking
where mails originate from, but requiring an explicit source helps to
expose bugs in the parser; such checks helped to identify the complications
described in \sectionref{discarding cleanup log lines} and
\sectionref{pickup logging after cleanup}.

Process scheduling can have a still more confusing effect: the child mail
will often be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  Unfortunately, it is not possible
to distinguish between child mails waiting to be tracked and other mails
whose origin is unknown, except for bounce notifications, as described in
\sectionref{identifying bounce notifications}.  Mails whose origin is
unknown can remain in the state tables indefinitely if their origin is not
determined at some point; they will cause a queueid clash if the queueid is
reused, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} cannot be expected to fully understand and reconstruct a mail
when some of that mail's log lines are missing.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; the log file will have log
lines for a mail whose origin is unknown.  Similarities exist to the
problem of identifying re-injected mails discussed in \sectionref{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line that identifies their
origin.  Heuristics must be used to identify bounce notifications:

@@

    \item The message-id must have the correct format and contain the
        queueid of the mail.  Although a mail sent from elsewhere could
        easily have the correct message-id format, the chance that the
        queueid in the message-id would correspond with the queueid of the
        mail is extremely small.

@@

If a mail is misclassified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably, misclassification of
this kind is a benefit rather than a drawback, because other mails
generated internally by Postfix will be handled correctly.  Postfix 2.3 and
subsequent versions log the creation of a bounce message, so this
complication does not arise in their log files.  The solution to this
complication will help with solving the complication in \sectionref{Bounce
notification mails delivered before their creation is logged}.

@@

Postfix does not log a message making the client's behaviour clear, so
heuristics are required to identify when a delivery attempt is aborted in
this way.  A list of all mails accepted during a connection is saved in the
connection's state table entry, and the accepted mails are examined when
the disconnection action is invoked.  Each accepted mail is checked for the
following:

@@

    \item If third and subsequent results exist, were those log lines
        processed by the \action{SAVE\_DATA} action?  Any log lines after
        the first two should be informational only.

@@

If all the checks above are successful the mail is assumed to be an aborted
delivery attempt and is removed from the state tables.  There will be no
further log lines for such mails, so without identifying and discarding
them they accumulate in the state table and will cause clashes if the
queueid is reused.  Such mails cannot be entered in the database because
the only data they contain is the client hostname and IP address, but the
database schema requires many more fields --- see \sectionref{connections
table} and \sectionref{results table}.  These heuristics are quite
restrictive, and appear to have little scope for false positives; any false
positives would cause a warning when the next log line for such a mail is
parsed.  False negatives are less likely to be detected: there may be
queueid clashes (and warnings) if mails remain in the state tables after
they should have been removed; if the queueid is not reused, the only way
to detect false negatives is to examine the state tables after each parsing
run.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are handled elsewhere in the parser, e.g.\
\sectionref{timeouts during data phase}.  Postfix does not log a message
saying the mail has been discarded, as should be expected by now.  The
checks to identify this happening are:

@@

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection log line and a mail acceptance log
        line, followed by one or more delivery attempt rejected log lines.

@@

If both checks are successful the mail is assumed to have been discarded by
Postfix when the client disconnected; \parsername{} will remove it from the
state tables.  There will be no further log lines for such mails, so if
\parsername{} does not deal with them immediately they accumulate in the
state table and will cause clashes if the queueid is reused.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes a timeout occurs or the
connection is lost during the DATA phase;\footnote{For the sake of brevity
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and \parsername{} needs to discard the data associated with that
mail.  It seems more appropriate to save the mail's data to the database,
but if a timeout occurs no data will be available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the \action{TIMEOUT} action does the following:

@@

    \item Record the timeout and data extracted from the log line in the
        connection's results.

    \item If no mails have been accepted over the connection, nothing needs
        to be done; the \action{TIMEOUT} action ends.

@@

            \item The timestamp of the log line preceding the timeout log
                line must be earlier than the timestamp of the last
                accepted delivery attempt, i.e.\ there have not been any
                rejections since then the delivery attempt was accepted.

@@

            \item The mail must not have a \daemon{qmgr} log line.

@@

        If all checks pass the mail will be discarded from the state tables
        and will not be entered in the database.  If one or more checks do
        not pass, \parsername{} assumes that the timeout happened after a
        rejected delivery attempt.  This assumption is not necessarily
        correct, because Postfix could have accepted an earlier recipient,
        rejected a later one, and continued to accept delivery of the mail
        for the first recipient.  In that case the timeout applies to the
        partially accepted mail, which will be discarded by Postfix and
        should be discarded by \parsername{}; however, this has not
        occurred in practice.  Processing timeouts is further complicated
        by the presence of out of order \daemon{cleanup} log lines: see
        \sectionref{discarding cleanup log lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail
because it did not have enough data to satisfy the database schema (see
\sectionref{timeouts during data phase}) and was not expected to have any
more log lines; parsing the \daemon{cleanup} log line causes the
\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
incorrect because the log line actually belongs to the mail that has just
been discarded; if the queueid is reused there will be a queueid clash,
otherwise the new mail will just remain in the state tables.

During the \action{TIMEOUT} action, if the mail's \daemon{cleanup} log line
is still pending, the \action{TIMEOUT} action updates the
\texttt{timeout\_queueids} state table, adding the queueid and the
timestamp from the log line.  To deal with this complication, the following
checks will be performed for each \daemon{cleanup} log line that is
processed:

@@

    \item Does the \texttt{timeout\_queueids} state table have an entry for
        the queueid in the log line?  If an entry is found it will be
        removed, regardless of whether the remaining criteria are
        satisfied.  If the queueid does not exist in the
        \texttt{timeout\_queueids} state table, the log line being
        processed cannot belong to a discarded mail.

    \item Has the queueid been reused yet, i.e.\ does it have an entry in
        the \texttt{queueids} state table?  If it has an entry in the
        \texttt{queueids} state table, the log line being processed belongs
        to that mail, not to a previously discarded mail.

@@

The \daemon{cleanup} log line must pass the checks above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
\action{CLEANUP\_PROCESSING} action blindly discarded the next
\daemon{cleanup} log line after a timeout it would sometimes be mistaken.

@@

When mail is submitted locally, \daemon{pickup} processes the new mail and
generates a log line.  Occasionally, this log line will occur later in the
log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
will find that a state table entry already exists for that queueid.
Normally when this happens a warning is generated by the \action{PICKUP}
action, but if the following conditions are met it is assumed that the log
lines were out of order:

@@

    \item The only program that has logged anything so far for the mail is
        \daemon{cleanup}.

@@

As always with heuristics, there may be circumstances in which these
heuristics match incorrectly, but none have been identified so far.  This
complication only seems to occur during periods of particularly heavy load,
so is most likely caused by process scheduling vagaries.

@@

    \item Occasionally a \daemon{smtpd} will exit with an error, so the
        active connection must be cleaned up, entered in the database if it
        has enough data, and deleted from the connections state table.

    \item Every Postfix process uses a watchdog timer that kills the
        process if it is not reset for a considerable period of time (five
        hours by default).  This safeguard prevents Postfix processes from
        running indefinitely and consuming resources if a failure causes
        them to enter a stuck state.

@@

The circumstances above account for all occasions where a \daemon{smtpd}
suddenly stops logging.  In addition to removing an active connection from
the state tables, the last accepted mail may need to be discarded, as
described in \sectionref{timeouts during data phase}; otherwise the queueid
state table is untouched.

@@

Occasionally, a log file will have out of order log lines that cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
logging after cleanup}.  In the \numberOFlogFILES{} log files used in
\sectionref{Evaluation} this problem occurs only five times in 60,721,709
log lines.  All five occurrences have the same characteristics: the
\daemon{local} log line showing delivery to a user's mailbox comes after
the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is complete.  This causes several problems: the data in
the state tables for the mail is not complete, so it cannot be entered into
the database; a new mail is created when the \daemon{local} log line is
processed and remains in the state tables; four warnings are issued per
pair of out of order log lines.

The \action{COMMIT} action examines the list of programs that have produced
log lines for each mail, checking the list against a table of known-good
Postfix component combinations.  If the mail's combination is found in the
table it can be entered in the database; if the combination is not found
entry must be postponed and the mail flagged for later entry.  The
\action{MAIL\_DELIVERED} action checks for that flag; if the log line is
has just processed has caused the mail to reach a valid combination then
entry of the mail into the database will proceed, otherwise it must be
postponed once more.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but those described in this section only occur
four times in the \numberOFlogFILES{} log files used in
\sectionref{Evaluation}.  The symptoms are the same as in
\sectionref{aborted delivery attempts}, except that the mail \textit{has\/}
a \daemon{cleanup} log line; nothing can be found in the log files to
explain why this mail does not have any further log lines.  The only way to
detect these mails is to periodically scan all mails in the state tables,
deleting any mails with the following characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted \parsername{} cannot determine the
recipient address or addresses.  This can occur when mail is accepted
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail before Postfix has had a chance to attempt delivery.  The
deleted mail's recipient address or addresses will not have been logged
yet, and the deleted mails will not have any more log lines.  A dummy
recipient address needs to be added by \parsername{}, because every mail is
required by the database schema (\sectionref{results table}) to have at
least one recipient.  When this complication occurs the log file will
typically show many instances of it, closely grouped.  Generally, this
problem arises because the administrator has deleted some mails from
Postfix's mail queue to stop a mail loop.

This lack of information cannot easily be overcome: it is simple to
configure Postfix to log a warning for every accepted recipient, but
Postfix will not yet have allocated a queueid for the mail when the warning
for the first recipient is logged, so the warning will be associated with
the connection rather than the accepted mail.  A queueid will be allocated
after Postfix accepts the MAIL FROM command if
\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
setting will cause disk IO for every delivery attempt, instead of just for
delivery attempts where recipients are accepted, and consequently a drastic
reduction in the performance of the mail server.

@@

\subsection[Bounce Notification Mails Delivered Before Their Creation Is
Logged]{Bounce Notification Mails Delivered Before \newline{} Their
Creation Is Logged}

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication, bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their origin.  The origin
of the mail is unknown when it is created, but when the time comes to enter
it into the database its origin is correctly identified by the heuristics
used to identify Postfix 2.2 bounce notifications, described in
\sectionref{identifying bounce notifications}.  To avoid incorrectly
creating a new mail when the out of order bounce notification log line is
processed, the \action{COMMIT} action maintains a cache of recently
committed bounce notification mails named \texttt{bounce\_queueids}, which
the \action{BOUNCE\_CREATED} action subsequently checks when processing the
bounce creation log line.  If the queueid exists in the cache, and its
start time is less than ten seconds before the timestamp of the bounce log
line, it is assumed that the bounce notification mail has already been
processed and the \action{BOUNCE\_CREATED} action does not create one.  If
the queueid exists in the cache it is removed, because it has either just
been used or the problem did not occur for the new mail.  Whether the
\action{BOUNCE\_CREATED} action creates a new mail or finds an existing
mail in the \texttt{queueids} state table (not the
\texttt{bounce\_queueids} cache), it flags the mail as having been seen by
the \action{BOUNCE\_CREATED} action; if this flag is present the
\action{COMMIT} action will not add the mail to the
\texttt{bounce\_queueids} cache.  This prevents a bounce notification log
line being incorrectly discarded if a queueid is reused within 10 seconds.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally,
mails that are in the process of being delivered will be deleted by the
administrator.  This results in the log lines from the delivery agent
(\daemon{local}, \daemon{smtp}, or \daemon{virtual}) appearing in the log
file \textit{after\/} the mail has been removed from the state tables and
saved in the database.  The \action{DELETE} action adds deleted mails to a
cache named \textit{postsuper\_deleted\_queueids}, which is checked by the
\action{MAIL\_DELIVERED} action, and the current log line discarded if the
following conditions are met:

@@

Sadly, this solution involves discarding some data, but the complication
only arises eight times in quick succession in one log file, which is not
in the \numberOFlogFILES{} log files used for evaluating the parser; if
this complication occurred more frequently it might be desirable to find
the mail in the database and add the log line to it.
